---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# matrixset

<!-- badges: start -->
[![R-CMD-check](https://github.com/pascalcroteau/matrixset/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pascalcroteau/matrixset/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

A `matrixset` a container of matrices, each having the same number of rows and
columns and the same dimnames. Moreover, each dimname must uniquely
identify elements.

While there is minimal support for `NULL` dimnames (and that is bound to
change at some point in the future), it is strongly recommended to
provide meaningful dimnames. One of the main reason for this is that
annotation is impossible with `NULL` dimnames.

In addition, as alluded above, a `matrixset` can store independent row
and column annotations. This meta information is stored, and available,
in the form of data frames - one for row information and one for column.
The annotation names are referred to as traits.

This latter feature makes `matrixset` especially attractive even if it
stores only a single matrix, because several methods have been
developped to manipulate `matrixset`s, accounting for annotations.

## Why a matrixset?

Many problems that `matrixset` can tackle could be solved via a `data.frame` and
more specifically using the `tidyverse` suite.

Two reasons for which you may want to use a `matrixset` instead are:

 * object size. The `data.frame` needed to store the same information as a
   `matrixset` can be significantly bigger
 * You actually need a matrix format, for example for running a PCA.

## Installation

You can install the development version of matrixset from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pascalcroteau/matrixset")
```

## Example

In addition to store multiple matrices that share the same attributes, a 
`matrixset` object's strength is it's annotation feature. You create an object
from existing matrices and annotation `data.frame`.

```{r, message=FALSE}
library(MASS)
library(tidyverse)
animals <- as.matrix(Animals)
head(animals)


animal_info <- MASS::Animals %>% 
  rownames_to_column("Animal") %>% 
  mutate(is_extinct = case_when(Animal %in% c("Dipliodocus", "Triceratops", "Brachiosaurus") ~ TRUE,
                                TRUE ~ FALSE),
         class = case_when(Animal %in% c("Mountain beaver", "Guinea pig", "Golden hamster", "Mouse", "Rabbit", "Rat") ~ "Rodent",
                           Animal %in% c("Potar monkey", "Gorilla", "Human", "Rhesus monkey", "Chimpanzee") ~ "Primate",
                           Animal %in% c("Cow", "Goat", "Giraffe", "Sheep") ~ "Ruminant",
                           Animal %in% c("Asian elephant", "African elephant") ~ "Elephantidae",
                           Animal %in% c("Grey wolf") ~ "Canine",
                           Animal %in% c("Cat", "Jaguar") ~ "Feline",
                           Animal %in% c("Donkey", "Horse") ~ "Equidae",
                           Animal == "Pig" ~ "Sus",
                           Animal == "Mole" ~ "Talpidae",
                           Animal == "Kangaroo" ~ "Macropodidae",
                           TRUE ~ "Dinosaurs")) %>% 
  select(-body, -brain)
animal_info
```

You can create the object and then do some operations.
```{r}
library(matrixset)
animals_ms <- matrixset(msr = animals, row_info = animal_info, row_key = "Animal")
animals_ms %>% 
    apply_row_dfl(rg = range(.i),
                  qt = quantile(.i, probs = c(.25, .75)))   


animals_ms %>% 
    row_group_by(class) %>% 
    apply_column_dfl(avr = mean)
```
